## Activity: RStudio

### Purpose

The purpose of this activity is to introduce the RStudio environment on SciServer for importing, exploring, filtering, and visualizing genomic data. 
Students will gain hands-on experience working with real bioinformatics output files (geNomad plasmid and virus summaries) and get a taste or foundational data analysis skills using R.

### Learning Objectives

By the end of this activity, students will be able to:

1. Data import and environment setup

- Import tabular data files into RStudio using functions such as read_tsv()
- Navigate and use the SciServer RStudio interface, including file upload and script execution

2. Data exploration and manipulation

- Explore datasets using basic R commands (e.g., viewing objects, inspecting column names)
- Sort and filter data using base R and tidyverse functions (e.g., sort(), filter(), %>%)
- Identify and subset biologically relevant data (e.g., contigs with AMR genes, non-provirus viral contigs)

3. Data Visualization and reproducibility

- Create and customize basic data visualizations using histogram plots in R (e.g., modifying titles and colors)
- Write, annotate, and organize reproducible R scripts for data analysis workflows

### Activity 1 – Import files to RStudio on SciServer

*Estimated time: 10 min*

#### Instructions

1. Download files.
    - Download locally two geNomad output files - plasmid summary and virus summary - from Galaxy history below, to your desktop. 
    - The files correspond to geNomad output for Zymo gut standard D6331 subset:
      - <https://usegalaxy.org/u/valerie-g/h/genomad-plasmid-virus-summary-zymo-gut-standard-d6331-subset>
1. Launch RStudio on SciServer.
    a. Login to SciServer.
      - <https://apps.sciserver.org/dashboard/>
    b. Navigate to SciServer ‘Compute’ (one of the cards on the bottom of the page).
    c. Click “Create container”.
    d. Give your container a name (eg. RStudio)
    e. In the “Compute Image” drop-down menu, select `Bioconductor 3.17 (RStudio)`
    f. Hit the green ‘Create’ button on the bottom.
    g. Click on the ‘RStudio’ Name to open RStudio window.
3. Upload files to RStudio on SciServer.
    a. Navigate to the `Files` pane located in the lower-right panel of the RStudio.
    b. Click the **'Upload'** button in the toolbar of the Files pane.
    c. Click **'Choose File'** to upload the two geNomad output files (you previously downloaded in Step 1 - uploading one at a time.
    d. You should now see two downloaded files in your ‘Files’ pane.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1e6d6Z51TJfpo6qrikIqnRwcJe6NTsxOj0tpJj7Fst8Y/edit?slide=id.p#slide=id.p")
```

4. Import files into RStudio.
    a. Use the top left window of RStudio - a blank R Script editor - to enter your R commands for importing files.  
    b. Enter a command to load R Package called ‘tidyverse’ by using **library()**:

```
library(tidyverse)
```

    c. Use the function **read_tsv()** in R to import files of interest.
        - See sample command below to import file called “Galaxy1-[geNomad-plasmid-summary].tabular” as a new object (variable) called e.g. ‘plasmids’.
        - After Typing your command, click **‘Run’**.

```
plasmids <- plasmids <- read_tsv("Galaxy1-[geNomad-plasmid-summary].tabular")
```

#### Questions

|1. Provide the command you used to import your **geNomad-plasmid-summary** file:|
| :--------- | 
| <br>       |

    
|2. Provide command you used to import your **geNomad-virus-summary** file:|
| :--------- | 
| <br>       |


### Activity 2 – Explore files in RStudio on SciServer

*Estimated time: 10 min*

#### Instructions

1. Use basic R commands to explore imported files in RStudio.
1. Use the top left window of RStudio - a blank R Script editor - to type and run your R commands - **you will be asked to provide your R Script code**.
    - Include all commands you used
    - Annotate all commands you used
    - See sample R Code below:

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1e6d6Z51TJfpo6qrikIqnRwcJe6NTsxOj0tpJj7Fst8Y/edit?slide=id.g3cf989fcd7c_0_11#slide=id.g3cf989fcd7c_0_11")
```

Your R Script for Activity 2 should include:

```
# Part 1: Plasmids


### Explore object

### Explore column names

### Sort file based on plasmid length, from high to low

### Filter file to only include contigs with detected antibiotic resistance genes (amr_genes)



# Part 2: Virus

### Explore object

### Explore column names

### Sort file based on plasmid length, from high to low

### Filter file to exclude Provirus
```

#### Questions

#### Part 1 - Explore geNomad plasmid summary file in RStudio

**1. View your imported plasmid file by typing in the name of your object - **plasmids**.

```Command: plasmids```

|Question: How many plasmid contigs (rows) are in your file?|
| :--------- | 
| Answer: |
| <br>       |


**2. View column names using function **colnames()**.

```Command: colnames(plasmids)```

|Question: Copy and paste column names below.|
| :--------- | 
| Answer: |
| <br>       |


**3. Sort plasmid contigs based on decreasing length, using function **sort()**.
    - Store as a new object called **“sorted”**; 
    - After sorting, to view the ‘sorted’ object, type ‘sorted’.

`Note: the **sort()** command in the example below is a function from the **R base package**. It is available in R without installing any special libraries.`

```
Your command will be the composite of following operations:
    - sorted <-: assigns the result of the operation to a new variable
    - sort (): the function used to do the sorting
    - plasmid$length: selects the “length” column from the plasmids object
    - Decreasing = TRUE: specifies that the sorting order should be descending rather than the default ascending.
```

Use these commands:

```
Command 1: sorted <- sort(plasmids$length, decreasing = TRUE)
Command 2: sorted
```

|Question: What is the largest contig length in a file? This should be the 1st number?|
| :--------- | 
| Answer: |
| <br>       |

4. Filter plasmid contigs using **filter()** function to retain only those that contain antimicrobial resistance (AMR) genes (amr_genes column); For this, you will need to remove the missing values (NA). Once filtered, view filtered contigs by calling on ‘filtered’ object.
    - **Store as a new object called “filtered”**; 
    - After filtering, to view the ‘filtered’ object, type **‘filtered’**.

`Note: the **filter()** function, used with the **%>%** pipe operator in the example below, is part of the **‘dplyr’ package - a component of package ‘tidyverse’** - which is a popular extension package in R, and is not a base R function. 

So, you can mix and match functions from different packages!`

```
Your command includes the following operations:
    - %>%: is the operator that sends the data into the next function
    - filter(): this function selects rows based on a condition
    - !=: is the “Not equal to” operator
```

Use these commands:

```
Command 1: filtered <- plasmids %>% filter (amr_genes != "NA")
Command 2: filtered
```

|Question: How many plasmids contigs had an AMR gene?|
| :--------- | 
| Answer: |
| <br>       |

#### Part 2 - Explore geNomad virus summary file in RStudio 

- Use commands you learned in Part 1 above, as a guide to explore virus summary file.

1. View your imported plasmid file by typing in the name of your object.

|1A: Type your command below.|
| :--------- | 
| |
| <br>       |

|1B: How many virus contigs (rows) are in your file?.|
| :--------- | 
| |
| <br>       |

2. View column names using function colnames().

|2A: Type your command below.|
| :--------- | 
| |
| <br>       |

|2B: Copy and paste column names below.|
| :--------- | 
| |
| <br>       |

3. Sort virus contigs based on decreasing length, using function **sort()**. Provide your commands and answer the question below.
    - Store as a new object called **“sortedV”**; 
    - After sorting, to view the ‘sorted’ object, type **‘sortedV’**.

|3A: Type your commands below:|
| :--------- | 
| Command 1:|
| Command 2: |
| <br>       |

|3B: What is the largest contig length in a file?|
| :--------- | 
| |
| <br>       |


4. Filter viral contigs using **filter()** function to all viral contigs EXCEPT those identified as “Provirus” (topology column); For this, you will need to eliminate the “Provirus”. 
    - Store as a new object called **“filteredV”**; 
    - After filtering, to view the ‘filtered’ object, type **‘filteredV’**.

|4A: Type your commands below:|
| :--------- | 
| Command 1:|
| Command 2: |
| <br>       |


|4B: How many viral contigs were NOT classified as Provirus?|
| :--------- | 
| |
| <br>       |


### Activity 3 – Plot and modify histogram of plasmid lengths

*Estimated time: 10 min*

#### Instructions

1. Use basic R commands to plot and modify histogram in RStudio
2. Continue to add this code to your R Script. 

#### Questions

#### Part 1 - Plot histogram of plasmid lengths

1. Plot histogram of plasmid contig lengths using function **hist()** and specifying the length column.
    - Your histogram will appear in the Plots tap (bottom right)

Use this command: 
```hist(plasmids$length)```

|1A: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |

2. Modify your histogram by changing the title to “Plasmid Length Distribution” 

Use this command: 
```hist (plasmids$length, main = "Plasmid Length Distribution")```

|2A: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |

3. Modify your histogram by changing histogram color to lightblue.

Use this command: 
```hist (plasmids$length, main = "Plasmid Length Distribution", col = "lightblue")```

|3A: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |

#### Part 2 - Plot histogram of virus lengths

1. Plot histogram of viral contig lengths using function hist() and specifying the length column
    - Your histogram will appear in the Plots tap (bottom right)

|1A: Type your command below: |
| :--------- | 
| |
| <br>       |


|1B: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |


2. Modify your histogram by changing the title to “Virus Length Distribution” 


|2A: Type your command below: |
| :--------- | 
| |
| <br>       |


|2B: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |

3. Modify your histogram by changing histogram color to 'lightgreen'.

|3A: Type your command below: |
| :--------- | 
| |
| <br>       |


|3B: Paste resulting plot below. |
| :--------- | 
| |
| <br>       |

#### Part 3 - Copy and paste your R Script below, including:

```
# Import files into R Studio

### Plasmids

### Virus


# Explore imported files

### Plasmids

### Virus



# Plot histogram of lengths

### Plasmids

### Virus
```


### Grading Criteria

- Download as Microsoft Word (.docx) and upload on Canvas

### Footnotes

**Resources**

- Google Doc

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Frederick Tan, Johns Hopkins University

Last Revised: March 2026




