## Activity: geNomad

### Purpose

Identify sequences of plasmids and viruses with [geNomad](https://doi.org/10.1038/s41587-023-01953-y). 
geNomad is one of the most modern plasmid and virus classification and annotation tools. 
geNomad uses a dataset of over 200,000 gene markers specific to chromosomes, plasmids or viruses to classify sequences as plasmids or viruses. 
Concurrently, geNomad uses a gene-independent and reference alignment-independent deep neural network approach that classifies sequences  as plasmids of viruses. 
With geNomad, using contigs as input you can:

- Identify sequences of plasmids and viruses 
- Functionally annotate plasmid and viral genes
- Taxonomically classify viral sequences

### Learning Objectives

1. Run a geNomad tool in Galaxy
1. Examine geNomad output (plasmids and virus)
1. Plot geNomad output (plasmid length and number of genes) in  RStudio on Galaxy

### Activity 1 – Run geNomad

*Estimated time: 10 min*

#### Instructions

1. Import the dataset corresponding to assembled contigs from PacBio-sequenced Zymo gut standard D6331 - a 3.4 Gb data subset assembled with Flye tool.
    - <https://usegalaxy.org/u/valerie-g/h/flye-assembly-zymo-gut-standard-d6331-subset>
1. Go to Tools and find geNomad
  a. Agree with geNomad licence
  b. Use assembled contigs (from Flye assembly) as input
  c. Under **filtering presets**, select **Manual settings** which correspond to geNomad's default setting.
  d. Click **Run Tool**

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1qSW-UU25kX9SPjfupKOzOjqRP6XeBVSJH-r1uVNaaKk/edit?slide=id.g3cd65ceba65_0_14#slide=id.g3cd65ceba65_0_14")
```

#### Questions

|1. Provide your Galaxy history with geNomad output:|
| :--------- | 
|             |

|2. Based on geNomad output file names, what information does geNomad provide?|
| :--------- | 
|             |

### Activity 2 – Explore geNomad output in Galaxy

*Estimated time: 10 min*

#### Instructions

1. Import two of eight geNomad output files for Zymo gut standard D6331 (Activity 1 output): 
    - <https://usegalaxy.org/u/valerie-g/h/genomad-plasmid-virus-summary-zymo-gut-standard-d6331-subset>
1. View geNomad output files to explore.

#### Questions

**1. Click on geNomad-plasmid-summary file** and answer the questions below:

|A. How many contigs were classified as plasmid-derived contigs?|
| :--------- | 
|             |
<br>

|B. Does every plasmid contig have a conjugation gene?|
| :--------- | 
|             |
<br>

|C. Does every plasmid contig have an amr_gene (anti-microbial resistance gene)?|
| :--------- | 
|             |
<br>

**2. Click on geNomad-virus-summary file** and answer the questions below:

|A. How many contigs were classified as virus-derived contigs?|
| :--------- | 
|             |
<br>

|B. Google search full taxonomy of the first viral contig in the file and record below 1 thing you learned about this virus.|
| :--------- | 
|             |
<br>

|C. What is the most common genomic context (topology) in which virus was identified in this data subset?|
| :--------- | 
|             |
<br>

### Activity 3 – Examine geNomad plasmids in RStudio

*Estimated time: 15 min*

#### Instructions

1. Launch RStudio tool in Galaxy
  - You don’t need to include input datasets with you RStudio launch - we will import data once RStudio is launched.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1qSW-UU25kX9SPjfupKOzOjqRP6XeBVSJH-r1uVNaaKk/edit?slide=id.g3cd91cb0fd0_0_4#slide=id.g3cd91cb0fd0_0_4")
```

2. Import data into RStudio
    a. **In your Galaxy history**, identify which Galaxy history number (folder) corresponds to the plasmid summary output file. 
      -  Lets assume dataset 41 in your Galaxy history is a plasmid summary file from geNomad and you want to read it into your RStudio.

    b. In your **RStudio console**, use function **gx_get()** to import (copy) a dataset of interest from Galaxy history, to RStudio session.

```
# Get Galaxy history #41|
(gx_get(41))
```

    c. In addition, you have to use a proper R function to read the file. To read tabular files, use function **read_tsv()**. To do so, you will first need to load an R package called **tidyverse**. Use the following pieces of code:

```
#Load tidyverse
library(tidyverse)

# Read a .tsv file
read_tsv(gx_get(41))
```

  d. Now that you have all pieces of code, import your tsv file as an object called, e.g. “plasmids” (or give it another convenient name of your choice).

```
#Final import command
plasmids <- read_tsv(gx_get(41))
```

  e. Once code is ready, type the 2 commands (to load tidyverse, and to import plasmid_summary tabular file) into your RStudio console:

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1qSW-UU25kX9SPjfupKOzOjqRP6XeBVSJH-r1uVNaaKk/edit?slide=id.g3cd91cb0fd0_0_53#slide=id.g3cd91cb0fd0_0_53")
```

3. Create an R script of the commands from step 2: 
  - Just like the best practice for wet-lab experiments is keeping a lab notebook, the best practice for computational experiments is keeping a notebook with all your code (commands) - e.g. having a record of your R script.
  - For best practice, you should annotate each block of code

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1qSW-UU25kX9SPjfupKOzOjqRP6XeBVSJH-r1uVNaaKk/edit?slide=id.g3cd91cb0fd0_0_20#slide=id.g3cd91cb0fd0_0_20")
```

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1qSW-UU25kX9SPjfupKOzOjqRP6XeBVSJH-r1uVNaaKk/edit?slide=id.g3cd91cb0fd0_0_75#slide=id.g3cd91cb0fd0_0_75")
```

#### Questions

#### Part 1 - Explore data in RStudio

1. Preview first rows of a file using function **head()**.

|Command: head(plasmids)|
| :--------- | 
|  Question:  Copy and paste output of command below: |
|  Answer:            |
<br>

2. View column names using function **colnames()**.

|Command: colnames(plasmids)|
| :--------- | 
|  Question:  Copy and paste output of command below: |
|  Answer:            |
<br>

3. Summarize each variable using funciton **summary()**.

|Command: summary(plasmids)|
| :--------- | 
|  Question:  Copy and paste output of command below: |
|  Answer:            |
<br>

4. Check the number of rows and columns - dimensions - of the file using function **dim()**.

|Command: dim(plasmids)|
| :--------- | 
|  Question:  Copy and paste output of command below: |
|  Answer:            |
<br>


#### Part 2 - Plot histogram of plasmid lengths

1. Plot histogram of plasmid lengths using function **hist()**.
  - Note, a dollar sign ($) in the command below indicates column

|Command: hist(plasmids$length)|
| :--------- | 
|  Question:  Copy and paste output of command below: |
|  Answer:            |
<br>

#### Part 3 - Plot histogram of plasmid gene density

1. Plot histogram of plasmid lengths using function **hist()**.
  - Note, a dollar sign ($) in the command below indicates column


|A. Type your command below:|
| :--------- | 
|  |
<br>

|B. Paste your output histogram below::|
| :--------- | 
|  |
<br>

### Grading Criteria

- Download as Microsoft Word (.docx) and upload on Canvas

### Footnotes

**Resources**

- Google Doc

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Frederick Tan, Johns Hopkins University

Last Revised: March 2026

