---
title: "CURE: Microbial Mysteries"
date: "`r format(Sys.time(), '%B %d, %Y')`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Description about Course/Book."
favicon: assets/dasl_favicon.ico
output:
    bookdown::gitbook:
      # config:
      #   edit: https://github.com/C-MOOR/C-MOOR_Template/edit/main/%s
      split_by: section
    bookdown::word_document2:
      toc: true
---

# About {-}

Welcome to Microbial Mysteries, a Course-based Undergraduate Research Experience (CURE) that takes you on a genomics adventure from soil to gut!

This CURE allows students to develop a project using genomics datasets from environments such as human gut and soil, and learn how organisms, high-throughput sequencing, and the scientific process are used in both basic and clinical research.


```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g344ad28629a_0_0#slide=id.g344ad28629a_0_0")
```

### Skills Level {- .unlisted}

::: {.notice}
_Genetics_  
**Novice**: Introduction to [central dogma of molecular biology](https://openstax.org/books/biology-2e/pages/15-1-the-genetic-code)

_Programming skills_  
**Novice**: No programming experience needed

_Research skills_ 
**Novice**: No scientific researdch experience needed
:::


### Learning Goals {-}

- **Navigate biological databases** - access the wealth of publically available scientific knowledge and data (e.g. NCBI).
- **Analyze Big Data** - e.g. plot and interpret results from next generation sequencing (NGS) experiment.
- **Gain skills in computational and bioinformatics data analysis** - e.g. web-based platforms or command-line bioingofmatics tools (e.g. Galaxy, R).
- **Practice the scientific process and critical thinking** -  learing to read and evaluate scientific literature, ask questions, identify avenues for research, design experiments, analyze and interpret data, and integrating results into the broader scientific discourse.
- **Engage in scientific research** - Apply knowledge and skills learned to complete a project using existing datasets to ask new questions, generate novel conclusions and create and present a Final Poster.


### C-MOOR Content Collection {- .unlisted}

```{r, echo = FALSE, results='asis'}
cow::borrow_chapter(
  doc_path = "about_c-moor.Rmd",
  repo_name = "C-MOOR/C-MOOR_Template"
)
```
-->


