# Microbial Genomes

Microbial genomes range in size and organization. Despite their small size and having fewer number of coding genes, bacteria are more taxonomically diverse, more abundant and more successful in colonizing new habitats - they are present in virtually every environment on earth!. This is in part due to bacteria having very 'busy' genomes with a high proportion of protein-coding functional genes and regions (and very little noncoding DNA and introns) which provide a source of innovation, novelty and adaptation e.g. by simply altering (mutating) functional genes [(P.C. Kirchberger et al, 2020)](https://doi.org/10.1146/annurev-micro-020518-115822). In contrast, eukaryotic genomes mostly consist of noncoding DNA, and a lot of mutations are bound to occur in those abundant noncoding regions.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g370690c0d46_0_0#slide=id.g370690c0d46_0_0")
```

In this **Microbial Genomes module**, we will learn about microbial genomes and microbial databases. In the **Pre-lab** activity, we will use *E. coli* to explore NIH genetic sequence database **GenBank**, bacterial databases such as **BV-BRC** and tree of life database **Lifemap**. For **Project** activity, we will continue microbial genome exploration using a published metagenome-assembled genomes or **MAGs**, which are microbial genomes assembled (reconstructed from) metagenome sequencing data. We will learn about the taxonomy of the MAG, its size, and assembly statistics and about its genes.

## Lecture: Microbial Genome

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1bnFhIIu6ZXSCjlrz5qWc7BZNz-7VfYWkKgbz3kL-ZNU/edit?slide=id.g35f391192_00#slide=id.g35f391192_00")
```

[Slides: Microbial Genomes](https://docs.google.com/presentation/d/1rj8kWK3-i2py65A0HyDTTXQFaQZ8MX7SRQK2YG0XWQg)

## Prelab: Microbial Genomes

### Purpose

Impress all the information that is freely available about well studied (and not so well studied) bacterial species. 
Start with *E. coli* as one of the best and longest studied before going into recently discovered bacteria (with less information).

### Learning Objectives

1. GenBank – Explore sequence database of publicly available DNA sequences.
2. Sequence Browser – Observe genome organization using graphical representation.
3. Bacterial database BV-BRC - Explore bacterial genes and their function.
4. Taxonomy Browser – Identify relationships between taxa.
5. Lifemap - Tree of Life viewer to visualize relationships between taxa.

### Activity 1 – GenBank

*Estimated time: 10 min*

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g305048ba785_0_0#slide=id.g305048ba785_0_0")
```

#### Instructions

1. Navigate to **GenBank** [https://www.ncbi.nlm.nih.gov/genbank](https://www.ncbi.nlm.nih.gov/genbank) and enter ‘E. coli strain K-12’ in the search bar (which should default to **Nucleotide**). 
    - *E. coli* strain K-12 is a laboratory strain commonly used in research. This will give you back a list of over 40,000 sequences. 
2. A famous substrain of *E. coli* strain K-12 is substrain MG1655. Let's narrow down our search and in the search bar  enter ‘Escherichia coli str. K-12 substr. MG1655’. 
    - This should give you a list of over 9,000 *E. coli* sequences.
3. Let's explore one specific version of the MG1655 substrain with NCBI Accession ID **NZ_CP169634**
    - In the search bar enter ‘NZ_CP169634’. 
4. This should take you to this link here: [https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634](https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634). 
    - Based on this page, answer the following questions.

#### Questions

|1. What is the size of this *E. coli* genome in bp and Kbp?|
|:--|
| <br> |

|2. Is the *E. coli* genome linear or circular?|
|:--|
| <br> |

|3. When you use  the NCBI Accession ID ‘NZ_CP169634’ in your search, how many entries did it give back? Based on this, how useful are NCBI Accession IDs?|
|:--|
| <br> |

### Activity 2 – Genomes, Genes, and Other Databases

#### Part 1 - Genomes

*Estimated time: 5 min*

#### Instructions

After entering Accession ID NZ_CP169634, on the right of the page under **Related Information**, click on the **Assembly** link to explore genome assembly information.

#### Questions

|1. How many chromosomes does *E. coli* have?|
|:--|
| <br> |

|2. What is the genome coverage of this sequenced genome?|
|:--|
| <br> |

|3. How many genes were annotated for this genome?|
|:--|
| <br> |

#### Part 2 - Genes

*Estimated time: 10 min*

#### Instructions

Go back to [https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634](https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634) and near the top of the page, find and click on **Graphics** to explore the genome browser.

a. Hover along one of many green vertical sticks
    - Click the `+` icon a few times to "Zoom In" to a genomic region for higher. resolution
b. Hover along one of many red vertical sticks
    - Zoom in until you see a difference between the green and red sticks.

#### Questions

|1. What do the ‘green sticks’ represent?|
|:--|
| <br> |

|2. What do the ‘red sticks’ represent?|
|:--|
| <br> |

|3. Record 5 genes you found present in *E. coli*.|
|:--|
| Gene example: ampC |
| Gene 1 |
| Gene 2 |
| Gene 3 |
| Gene 4 |
| Gene 5 |

#### Part 3 - Databases

*Estimated time: 10 min*

#### Instructions

To learn more about the genes of interest and their function scientists often use specialized databases. One such bacterial database is BV-BRC [https://www.bv-brc.org](https://www.bv-brc.org). Use BV-BRC to find information about some of the *E. coli* genes.

For the 2 genes below and one gene of your choice from the activity above:

a. Expand **"Searches"** in the top menu bar and select **“Pathways”**. 
b. Type the gene name into **"Keyword"** and press `<ENTER>`. The resulting output may contain gene entries for many different bacterial genomes, and *E. coli* may be one of them. 
c. Check one of the boxes corresponding to a specific **Genome Name** to learn more about your gene, and answer the following questions.

#### Questions

Note that some genes of interest will not be found in the BV-BRC database.  Rather than reporting "No results found", pick another gene.

|1. For genes below pick an organism (Genome Name).|
|:--|:--| 
|**Gene ID** | **Genome Name** |
| Gene 1: ampC| Escherichia coli 07798 | 
| Gene 2: mgtA| |
| Gene 3: Your gene| |

|2. For the 3 genes above, record gene Product / Description.|
|:--|:--| 
|**Gene ID** | **Product / Description** |
| Gene 1: ampC| Beta-lactamase | 
| Gene 2: | |
| Gene 3: | |

|3. For the 3 genes above, record Pathway Name (relates to function).|
|:--|:--|
|**Gene ID** | **Pathway Name** |
| Gene 1: ampC| beta-Lactam resistance | 
| Gene 2: | |
| Gene 3: | |

### Activity 3 -  Taxonomy and Tree of Life

*Estimated time: 25 min*

#### Part 1

#### Instructions

1. Go back to NCBI enter the *E. coli* accession number again [https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634](https://www.ncbi.nlm.nih.gov/nuccore/NZ_CP169634). 
2. Under **Related Information** on the right, click on **Taxonomy** and then on the provided link for the *E. coli*. 
3. Find taxonomic lineage information labeled as **"Lineage(full)"**. Full lineage information contains seven core taxonomy ranks: Kingdom, Phylum, Class, Order, Family, Genus and Species, plus any additional classification ranks. To see which taxonomic rank a name refers to you can simply hover over the lineage name.

#### Questions

|1. Record the seven core taxonomy ranks below.|
|:--|:--|
| Kingdom: | |
| Phylum:| |
| Class:| |
| Order:| | 
| Family:| |
| Genus:| |
| Species:| |

|2. Use any search engine to learn 2-3 facts about phylum Pseudomonadota. What did you learn?|
|:--|
| Fact example: “Pseudomonadota” is also known as “Proteobacteria”.|
| Fact 1:|
| Fact 2:|
| Fact 3:|

#### Part 2

#### Instructions

1. As with the BV-BRC database above, we can use another database called **Lifemap** to visually explore *E. coli* in the context of the tree of life. Go to [https://lifemap-ncbi.univ-lyon1.fr](https://lifemap-ncbi.univ-lyon1.fr), click "Start Exploring", click "Search a taxa" (magnifying glass icon) in the left hand menu, type "E. coli", and click the "Species" results.
2. On the tree map, the green dot and circle will indicate *E. coli*. Use plus and minus tabs to zoom in and out and visualize *E. coli* relative to other organisms on the map. 

#### Questions

|1. How many Domains of life are there?|
|:--|
| <br> |

|2. Zoom into and find nodes for the *E. coli* Genus, Family, Order, Class and Phylum. Do they match what you found in the NCBI Taxonomy Browser for Lineage?|
|:--|
| <br> |

|3. What are some other members of the Genus (Escherichia) to which *E. coli* belongs?|
|:--|
| <br> |

|4. What are some other members of the Family (Enterobacteriaceae) to which *E. coli* belongs?|
|:--|
| <br> |

|5. What are some other members of the Order (Enterobacterales) to which *E. coli* belongs?|
|:--|
| <br> |

|6. What are some other members of the Class (Gammaproteobacteria) to which *E. coli* belongs?|
|:--|
| <br> |

|7. What are some other members of the Phylum (Pseudomonadota) to which *E. coli* belongs?|
|:--|
| <br> |

### Grading Criteria

- Download as Microsoft Word (.docx) and upload on Canvas

### Footnotes

**Resources**

- [Google Doc](https://docs.google.com/document/d/1M10YdzSMXZJv80_p0diK9rco3-N8yU4C)

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Frederick Tan, Johns Hopkins University

Last Revised: January 2026

## Discussion: Microbial Genomes

### Activity 

*Estimated time: 25 min*

#### Instructions

1. Form new groups of four.
    - [docs.google.com/spreadsheets/d/1yDkZQCPRKdorJeL7uqHe7zD33Yab34chfNV4Bk6_QLo](https://docs.google.com/spreadsheets/d/1yDkZQCPRKdorJeL7uqHe7zD33Yab34chfNV4Bk6_QLo)
2. Pair up into groups (10 min).
    - Discuss -- Each group member briefly describes answers to prelab assignment
    - Summarize -- Identify best answer and add to slidedeck
        - [docs.google.com/presentation/d/1PRJPnOWbV6SMH3ph3uenEY0gJE6LeOA-EEbhd_lPde8](https://docs.google.com/presentation/d/1PRJPnOWbV6SMH3ph3uenEY0gJE6LeOA-EEbhd_lPde8)
3. Share group discussion (2 min each group).

### Footnotes

**Resources**

- [Google Doc](https://docs.google.com/document/d/1oRrYkD3q8nWTUtrs6w0ZgMYYfyMTDoWJ)

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Frederick Tan, Johns Hopkins University

Last Revised: January 2026

## Project: Microbial Genomes

### Purpose

Explore information about new bacterial MAGs from Xue, et al. 2023 Scientific Data ([pubmed.gov/37563185](https://pubmed.gov/37563185)).

### Learning Objectives

1. Explore and better understand MAGs and contigs by following up one MAG and one contig in NCBI.
2. Utilize knowledge learned by navigating MAG information using previously learned tools such as GenBank, Sequence Browser, BV-BRC, Taxonomy Browser, and Lifemap. 
3. Go deeper and see what you can find on the organism from your MAG of choice and a gene from a MAG of choice using any tools available to you (Pubmed, Google, AI, BV-BRC, other databases). 

### Activity 1 – MAGs and Taxonomy

*Estimated time: 40 min*

#### Part 1 - Explore a MAG

#### Instructions 

1. In the research study by [Xue, et al. 2023](https://pubmed.gov/37563185), 17 of 103 uncovered MAGs from ballast water or sediment were of very high quality and completeness. You will follow up on one of them!
1. Choose **one of these 17 MAGs** (see their GenBank IDs **below**) to follow up in this activity.
1. Use GenBank [ncbi.nlm.nih.gov/nucleotide](https://www.ncbi.nlm.nih.gov/nucleotide/) to answer questions. 

|GenBank IDs for 17 high-quality MAGs| |
|:-------------------|:--------------------|
| 1. GCF_030147545.1 | 10. GCF_030149085.1 |
| 2. GCF_030148435.1 | 11. GCF_030149145.1 |
| 3. GCF_030149225.1 | 12. GCF_030148125.1 |
| 4. GCF_030148195.1 | 13. GCF_030149425.1 |
| 5. GCF_030148855.1 | 14. GCF_030149235.1 |
| 6. GCF_030148245.1 | 15. GCF_030148515.1 |
| 7. GCF_030149045.1 | 16. GCF_030147875.1 |
| 8. GCF_030148385.1 | 17. GCF_030149465.1 |
| 9. GCF_030147715.1 |                     |

#### Questions

|1. Record the GenBank ID for one of the 17 MAGs that you will research.| 
|:--|
| <br> |

|2. Record the GENOME name for the MAG assembly associated with the GenBank ID you entered.| 
|:--|
| <br> | 

Hints:

a. In GenBank [ncbi.nlm.nih.gov/nucleotide](https://www.ncbi.nlm.nih.gov/nucleotide/), under the Nucleotide search tab **enter** the GenBank number of your MAG (e.g. GCF_030147545.1) and click **Search**. 

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g305048ba785_0_77#slide=id.g305048ba785_0_77")
```

b. For that MAG (e.g. 1st MAG with GenBank ID GCF_030147545.1), the GENOME name is "Alcanivorax sp. genome ASM3014754v1":

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g357b6795023_0_2#slide=id.g357b6795023_0_2")
```

For the following questions, click on the GENOME name of your MAG to **explore genome assembly summary** information.

|3. What is the Taxon of your MAG?| 
|:--|
| <br> | 

|4. What is the genome size of your MAG?| 
|:--|
| <br> | 

|5. How many contigs contributed to your MAG assembly?| 
|:--|
| <br> | 

|6. How many genes were annotated?| 
|:--|
| <br> | 

|7. How comparable is your MAG genome to the E. coli genome from the microbial-genomes-pre-lab, in terms of genome size and number of genes?| 
|:--|
| <br> | 

|8. As you can see, MAGs are made up of Contigs. Based on your lecture and reading material, and this exercise, in your own words define MAGs and Contigs below.| 
|:--|
| MAGs:   |
| Contigs:|

#### Part 2 - Explore a Contig

#### Instructions

1. MAGs are made up of Contigs. To see which contigs make up your MAG, go back to GenBank [ncbi.nlm.nih.gov/nucleotide](https://www.ncbi.nlm.nih.gov/nucleotide/), and search for your MAG using its GenBank ID (e.g.GCF_030147545.1). 
    - For example, you will see that MAG GCF_030147545.1 is composed of 41 contigs.

```{r, echo=FALSE, fig.alt=''}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g357f32d72ad_0_0#slide=id.g357f32d72ad_0_0")
```

#### Questions

|1. Choose a contig of reasonably large size (> 75 kb), click on the contig, then find and record below the Contig’s ID (Accession number/GenBank ID) for further examination.|
|:--|
|Tip: You can sort the entries by length via’ Sort by Sequence Length’ on top!|
|<br>|

|2. Record 7 core taxonomy ranks for your Contig.| |
|:--|:--|
| Kingdom: | |
| Phylum:| |
| Class:| |
| Order:| | 
| Family:| |
| Genus:| |
| Species:| |

Hints:

a. For the contig you chose above, under **Related Information** on the right, click on **Taxonomy** and **then click again on the provided link**.
a. Find Lineage information. Full Lineage information contains 7 core taxonomy ranks: Kingdom, Phylum, Class, Order, Family, Genus and Species, plus any additional classification ranks. To just get the 7 core lineage names, click on Lineage link for the abbreviated Lineage, or, simply hover over lineage names. 

#### Part 3 - Visualize contig in a tree of life

#### Instructions

1. Use Lifemap to visually explore the contig taxonomy in the context of the tree of life. Go to [lifemap-ncbi.univ-lyon1.fr](https://lifemap-ncbi.univ-lyon1.fr) and **enter the lowest taxonomy rank observed for your contig** (most likely the species or genus level, but can also correspond to order or family).
1. On the tree map, use plus and minus tabs to zoom in and out and visualize your Contig entry relative to other organisms on the map. Zoom in and find nodes corresponding to the higher taxonomic ranks. For example, if your contig corresponds to genus level classification, you will not be able to identify species level information, but you will be able to identify the corresponding Family, Order, Class and Phylum. 

#### Questions

|1. Record 7 core taxonomy ranks for your Contig.| |
|:--|:--|
| Kingdom: | |
| Phylum:| |
| Class:| |
| Order:| | 
| Family:| |
| Genus:| |
| Species:| |

|2. What are some other members of the Genus to which your Contig belongs?|
|:--|
|<br>|

|3. What are some other members of the Family to which your Contig belongs?|
|:--|
|<br>|

|4. What are some other members of the Order to which your Contig belongs?|
|:--|
|<br>|

|5. What are some other members of the Class to which your Contig belongs?|
|:--|
|<br>|

|6. What are some other members of the Phylum to which your Contig belongs?|
|:--|
|<br>|

### Activity 2 – Genomes, Genes, and Databases

*Estimated time: 20 min*

#### Instructions

1. In GenBank [ncbi.nlm.nih.gov/nucleotide](https://www.ncbi.nlm.nih.gov/nucleotide/), for the contig you chose in Activity 1 click on **Graphics** to explore the genome browser and the genes. 

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g357f32d72ad_0_15#slide=id.g357f32d72ad_0_15")
```

2. **Select genes of interest**. A lot of genes found in the Contigs will have no ‘familiar’ short symbol and instead have a long alphabetical-and-numerical- name. Such genes are either uncharacterized, hypothetical or have functional or structural similarity to known genes/proteins, but have not been confirmed. However, some Contig genes will be annotated with a ‘familiar’ short gene symbol, matching known genes. Please use the genes with the short symbols for this activity, since the ‘other’ genes will not be found in the databases.

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Image test"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1hYKF7Ss3vJ8rrUIH7ByNh1BUlRa2fhsJhq8MXzEowCc/edit?slide=id.g357f32d72ad_0_31#slide=id.g357f32d72ad_0_31")
```

3. For your genes of interest, use **BV-BRC** [bv-brc.org](https://www.bv-brc.org) to find more information.

#### Questions

|1. Record 3 genes of interest for your Contig.| |
|:--|:--|
| Gene1: | |
| Gene2:| |
| Gene3:| |

|2. Record one organism (Genome Name) for which the gene of interest is available.| |
|:--|:--|
|**Gene**| **Genome Name**|
| Gene1: | |
| Gene2:| |
| Gene3:| |

Hint:

In the [bv-brc.org](https://www.bv-brc.org) **Search** space:

a. From a dropdown menu select **Pathways**, 
a. Type in the gene name and **click enter**. This may result in a lot of gene entries for different organisms, 
a. Note one host species/strain (**Genome Name**).

|3. Record gene Product associated with your genes of interest.| |
|:--|:--|
|**Gene**| **Product**|
| Gene1: | |
| Gene2:| |
| Gene3:| |

|4. Record Pathway Name associated with your genes of interest.| |
|:--|:--|
|**Gene**| **Pathway Name**|
| Gene1: | |
| Gene2:| |
| Gene3:| |

### Activity 3 – Go Deeper

*Estimated time: 30 min*

#### Instructions

Use any tools available at your disposal to follow up on your MAG and gene of interest from activities 1 and 2 above. Some suggested tools include PubMed, Google, AI, [MBGD](https://mbgd.nibb.ac.jp), BV-BRC, [BacDive](https://bacdive.dsmz.de).

#### Questions

1. For the **taxa** you identified for your chosen MAG in activity 1, what can you learn about this organism (species or genus for example) in 15 minutes using any tools at your disposal? 

|a. What did you learn?|
|:--|
|   |


|b. What tools did you use?|
|:--|
|   |

2. For one of the **genes** you identified for your chosen contig in activity 2, what can you learn about this gene in 15 minutes using any tools at your disposal?

|a. What did you learn?|
|:--|
|   |

|b. What tools did you use?|
|:--|
|   |

### Grading Criteria

- Download as Microsoft Word (.docx) and upload on Canvas

### Footnotes

**Resources**

- [Google Doc](https://docs.google.com/document/d/1JFnxV-bMo_Zr0q-a3tuYXVpZrBY7cN0C)

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Gauri Paul, Clovis Community College
- Frederick Tan, Johns Hopkins University

Last Revised: January 2026

## Presentation: Microbial Genomes

### Activity

*Estimated time: 25 min*

#### Instructions

1. Open the “Microbial Genomes Presentation” slidedeck
    - [docs.google.com/presentation/d/1rp5AYFGnzs2z8r2nSWf_JMDsLqLcCc0qigZ5mMe84Qs](https://docs.google.com/presentation/d/1rp5AYFGnzs2z8r2nSWf_JMDsLqLcCc0qigZ5mMe84Qs)
1. Create two slides to present your findings
    - Describe the MAG you explored (Activity 1 and 3)
    - Describe some of its genes (Activity 2 and 3)
1. Iterate your slides being sure to:
    - Add bullet points on key details you understand (Notice) and questions you have (Wonder)
    - Create slide titles that summarize your main takeaway
    - Include one or more plots or images that support your points
1. Present to the class (2 min each group)

### Footnotes

**Resources**

- [Google Doc](https://docs.google.com/document/d/1blRIoHFfEtwuwpj_T6UPXkCT4-jzRJur)

**Contributions and Affiliations**

- Valeriya Gaysinskaya, Johns Hopkins University
- Frederick Tan, Johns Hopkins University

Last Revised: February 2026
